M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2011-12-18 notes/work/ch01.tex
Work on thesis.

\section{Physics}
Electrokinetic phenomena is described by the equations described below.
An ion-exchanger is described by the chemical kinetics at its boundary.
The experiment consists of applying an electic field and measuring the drift velocity of the particles.

@@

We consider a spherical particle. At its surface ($r=1$ with normal $\bn = \brhat$), 
we have zero slip, anion impermeability, cation selectivity and particle's conductance:
\begin{eqnarray}
\bv & = & \bzero \\
\bn \cdot \bj_- & = & 0 \\
c_+ & = & \gamma \\
\varphi & = & \cV
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
most of the fluid remains electroneutral ($c_+ \approx c_-$), 
except of a thin boundary layer of width $O(\delta)$ 
, also known as ``Debye layer'', around the ion exchanger.

@@

In order to match bulk $O(1)$ outer radial ionic flux, 
the $O(\delta^{-1})$ term of the inner ionic radial flux $j_\pm(\rho, \theta)$ must vanish:
\begin{eqnarray}
  -\deriv{c_\pm}{\rho} \mp c_\pm \deriv{\varphi}{\rho} &=& 0 \\
  \deriv{\varphi}{\rho} \pm\frac{1}{c_\pm}\deriv{c_\pm}{\rho} &=& 0 \\
  \pars{\varPhi - \varphi} \pm\pars{\log C_\pm - \log c_\pm} &=& 0 \\
  \varPhi - \varphi \pm \log \frac{C_\pm}{c_\pm} &=& 0
\end{eqnarray}
This relation can be rewritten as Boltzmann distribution of ionic concentration,
since $C_\pm|_{\rho\rightarrow\infty} = C$:
\begin{eqnarray}
c_\pm = C \exp\left[\pm(\varPhi - \varphi)\right]
\end{eqnarray}

@@

Denote zeta potential as the voltage drop between the particle surface and the end of
the boundary layer, $\zeta = \cV - \varPhi$. 
Without loss of generality, 
we assume that $\cV = -\log\gamma$, thus $\zeta = -\log\gamma-\varPhi$:
\begin{eqnarray}
\varPhi + \log C &=& 0
\end{eqnarray}

@@

The leading-order electric potential in the Debye layer:
\begin{eqnarray}  
\deriv{^2 \varphi}{\rho^2} = -q = -\frac{c_+ - c_-}{2} = 
-C\frac{\exp\left[+(\varPhi - \varphi)\right] - \exp\left[-(\varPhi - \varphi)\right]}{2} = 
C \sinh(\varphi - \varPhi)
\end{eqnarray}
Denote $\psi = \varphi - \varPhi$, so that $\psi(0) = \cV - \varPhi = \zeta$ and 
$\psi(\rho\rightarrow\infty) = 0$.
\begin{eqnarray}  
\deriv{^2 \psi}{\rho^2} &=& C \sinh(\psi) = 2 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} 
\\
2\deriv{\psi}{\rho} \deriv{^2 \psi}{\rho^2} &=& 
4 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} \deriv{\psi}{\rho} 
\\
\deriv{}{\rho} \pars{\deriv{\psi}{\rho}}^2 &=& 
\deriv{}{\rho} \pars{2 \sqrt{C} \sinh{\frac{\psi}{2}}}^2 
\end{eqnarray}

Integrating with respect to boundary condtions yields:
\begin{eqnarray}  
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} \\  
\pars{4\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}}^{-1}\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\frac{\pars{\tanh\frac{\psi}{4}}^{-1}}{4\cosh^2\frac{\psi}{4}}
\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\deriv{}{\rho}\pars{\log\tanh\frac{\psi}{4}} &=& \deriv{}{\rho}\pars{-\rho \sqrt{C}} 
\end{eqnarray}

Integrating with respect to boundary condtions yields:
\begin{eqnarray}
\log\tanh\frac{\psi}{4} - \log\tanh\frac{\zeta}{4} &=& -\rho \sqrt{C} \\
\psi = \varphi - \varPhi &=& 
4 \tanh^{-1}\pars{\tanh{\frac{\zeta}{4}} \cdot \exp\pars{-\rho \sqrt{C}}}
\end{eqnarray}

Leading-order radial momentum balance in the Debye layer yields:
\begin{eqnarray} 
- \deriv{p}{\rho} + \deriv{\varphi}{\rho} \deriv{^2\varphi}{\rho^2} &=& 0
\end{eqnarray}
One integration yields:
\begin{eqnarray} 
p &=& \frac{1}{2}\pars{\deriv{\varphi}{\rho}}^2 = \frac{1}{2}\pars{\deriv{\psi}{\rho}}^2
\end{eqnarray}

Leading-order tangential momentum balance in the Debye layer yields:
\begin{eqnarray} 
0 &=& \deriv{^2 v_\theta}{\rho^2} - \deriv{p}{\theta} 
 + \deriv{\varphi}{\theta} \deriv{^2\varphi}{\rho^2} 
\\
\deriv{^2 v_\theta}{\rho^2} &=& \frac{1}{2}\deriv{}{\theta}\pars{\deriv{\psi}{\rho}}^2
- \deriv{(\psi + \varPhi)}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{2C \sinh^2\frac{\psi}{2}}
- \deriv{\psi}{\theta} C \sinh\psi
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{C \pars{\cosh\psi - 1}}
- C \sinh\psi \deriv{\psi}{\theta}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{C}{\theta} \pars{\cosh\psi - 1}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = \deriv{C}{\theta} \pars{2 \sinh^2 \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \pars{\half \sinh \frac{\psi}{2} \deriv{\psi}{\rho}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \deriv{}{\rho}\pars{\cosh \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\end{eqnarray}
By integrating from specific $\rho$ to $\rho \rightarrow \infty$ (where $\psi = 0$):
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} 
\pars{\cosh \frac{\psi}{2} - 1} - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
= -\deriv{C}{\theta} \frac{4}{\sqrt{C}} \sinh^2 \frac{\psi}{4} 
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
\end{eqnarray}
Note that:
\begin{eqnarray}
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} = 
                       -4\sqrt{C}\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}
\\
-4\sinh\frac{\psi}{4} &=& \frac{1}{\sqrt{C} \cosh\frac{\psi}{4}} \deriv{\psi}{\rho}
\end{eqnarray}
Thus:
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& 
  \pars{\deriv{C}{\theta} \frac{1}{C}} \cdot
  \pars{\frac{\sinh\frac{\psi}{4}}{\cosh\frac{\psi}{4}} \deriv{\psi}{\rho}}
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}
=  4\deriv{}{\rho} \pars{\log\cosh\frac{\psi}{4}} \deriv{}{\theta} \log C
  - \deriv{\psi}{\rho} \deriv{\varPhi}{\theta}
\\
\deriv{v_\theta}{\rho} &=& 
\deriv{}{\rho} \pars{ 4\log\cosh\frac{\psi}{4} \deriv{}{\theta} \log C
  - \psi \deriv{\varPhi}{\theta} }
\end{eqnarray}

By integrating from $\rho = 0$ to $\rho \rightarrow \infty$, 
we have Dukhin-Derjaguin slip formula for tangential velocity component $V_\theta$ 
(where the radial component is $V_r = 0$):
\begin{eqnarray} 
V_\theta &=& \zeta \cdot \deriv{\varPhi}{\theta} -
      4\log\cosh \frac{\zeta}{4} \cdot \deriv{}{\theta} \log C 
\\
&=& \zeta \cdot \deriv{\varPhi}{\theta} + 
      2\log\pars{1 - \tanh^2 \frac{\zeta}{4}} \cdot \deriv{}{\theta} \log C 
\end{eqnarray}
where $\zeta = \cV - \varPhi$.

Since $V_r = 0$ on the boundary, the slip condition can be written using surface
gradient operator:
\begin{eqnarray}
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C \\
\bnabla_\mathcal{S} &=& \pars{\tI - \bn\bn} \bnabla
\end{eqnarray}

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI \\
\bnabla \cdot \tT &=& \bzero
\end{eqnarray}
The total force acting on the particle is computed by the following surface integral:
\begin{eqnarray}
\bF &=& \oint_\mathcal{S} \tT \cdot \bnhat  dA 
\end{eqnarray}
In steady state (constant particle drift velocity $\cU$), it should be zero: $\bF = \bzero$.

>>>>

\section{Physics}
\subsection{Introduction}
Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{kirby2010book}. When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
\begin{figure}[htbp]
\begin{framed}
    \begin{center}
        \includegraphics[width=0.25\textwidth]
            {figs/ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer and (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        Zeta potential is defined as the voltage drop on (2).}
        \label{fig:EDL}
    \end{center}
\end{framed}
\end{figure}
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.

The partial differential equations that descibe the system dynamics
are coupled and non-linear. 
In general, they have no analytic solution. Moreover,
any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle's size. A
closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects for many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{pumera2010electrochemically,paxton2004catalytic,howse2007self}.
However, since the analytic model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver has to be developed.

@@

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we have zero slip, anion impermeability, cation selectivity 
and particle's conductance:
\begin{eqnarray}
\bv & = & \bzero \\
\bn \cdot \bj_- & = & 0 \\
c_+ & = & \gamma \\
\varphi & = & \cV
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
most of the fluid remains electroneutral ($c_+ \approx c_-$), 
except of a thin boundary layer of width $O(\delta)$, 
also known as ``Debye layer'', around the ion exchanger.

@@

In order to match bulk $O(1)$ outer radial ionic flux, 
the $O(\delta^{-1})$ term of the inner ionic radial flux $j_\pm(\rho, \theta)$ must vanish:
\begin{eqnarray}
  -\deriv{c_\pm}{\rho} \mp c_\pm \deriv{\varphi}{\rho} &=& 0 \\
  \deriv{\varphi}{\rho} \pm\frac{1}{c_\pm}\deriv{c_\pm}{\rho} &=& 0 \\
  \pars{\varPhi - \varphi} \pm\pars{\log C_\pm - \log c_\pm} &=& 0 \\
  \varPhi - \varphi \pm \log \frac{C_\pm}{c_\pm} &=& 0
\end{eqnarray}
This relation can be rewritten as Boltzmann distribution of ionic concentration:
\begin{eqnarray}
\lim_{\rho\rightarrow\infty} c_\pm &=& C \\
c_\pm &=& C \exp\left[\pm(\varPhi - \varphi)\right]
\end{eqnarray}

@@

Denote zeta potential as the voltage drop between the particle surface and the end of
the boundary layer, $\zeta = \cV - \varPhi$.

Without loss of generality, 
we assume that $\cV = -\log\gamma$, thus $\zeta = -\log\gamma-\varPhi$:
\begin{eqnarray}
\varPhi + \log C &=& 0
\end{eqnarray}

@@

Using the derivation at Appendix \ref{append:slip} and that $V_r = 0$ 
on the boundary, the slip condition can be written using surface
gradient operator:
\begin{eqnarray}
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C \\
\bnabla_\mathcal{S} &=& \pars{\tI - \bn\bn} \bnabla
\end{eqnarray}

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI \\
\bnabla \cdot \tT &=& \bzero
\end{eqnarray}
The total force acting on the particle is computed by the following surface integral:
\begin{eqnarray}
\bF &=& \oint_\mathcal{S} \tT \cdot \bnhat  dA 
\end{eqnarray}
In steady state (constant particle drift velocity $\cU$), it should be zero: 
\begin{eqnarray}
\bF = \bzero
\end{eqnarray}

### 2011-12-18 notes/work/ch02.tex
Work on thesis.

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Since the system has axial symmetry, the operators are written in 
spherical coordinates $(r,\theta,\phi)$:
\begin{eqnarray}
\brc{l}{
x = r \sin \theta \cos \phi \\
y = r \sin \theta \sin \phi \\
z = r \cos \theta} 
\end{eqnarray}
Due to axial symmetry, 
so the scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}
\end{eqnarray}
The scalar laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}
\end{eqnarray}
The vector gradient can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat
\end{eqnarray}

@@

\subsubsection{Grids choice}
In order to discretize the differential operators, 
a regular grid of size $n_r \times n_\theta$ is used:
\begin{eqnarray}
(r_i,\theta_j) &\in& [1, \infty) \times [0,\pi] \\ 
r_i &=& (1+\Delta_r)^i \\
\theta_j &=& \Delta_\theta \cdot j
\end{eqnarray}
where a logarithmic grid is used for $r$ and an uniform one is used for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$. Thus:
\begin{eqnarray}
\Delta_r &=& \pars{R_{max}} ^ \frac{1}{n_r} - 1 \\
\Delta_\theta &=& \frac{\pi}{n_\theta}
\end{eqnarray}

@@

Using an algebra of operators, the composition of operators is well defined:
\begin{eqnarray}
(\cO_1 \circ \cO_2)(\bx) &=& \cO_1(\cO_2(\bx))
\end{eqnarray}
Note that the differentiation of composed operators is computed by using the chain rule:
\begin{eqnarray}
\bnabla(\cO_1 \circ \cO_2)(\bx) &=& \bnabla \cO_1(\cO_2(\bx)) \cdot \bnabla \cO_2(\bx)
\end{eqnarray}

@@

The extended variables $\tilde X$ (with ``ghost'' points values 
from boundary conditions) are be computed, by applying the following non-linear operators:
\begin{eqnarray}
\tilde{\varPhi} &=& \cB_\varPhi(\varPhi, C; \beta) \\
\tilde{C} &=& \cB_C(C, \varPhi) \\
\tilde{V_r} &=& \cB_{V_r}(V_r; \cU) \\
\tilde{V_\theta} &=& \cB_{V_\theta}(V_\theta, C, \varPhi; \cU, \gamma) \\
\tilde{P} &=& P
\end{eqnarray}
The extended system variable can be written as $\tilde{\bx} = \cB(\bx)$, 
where $\cB$ is the ``ghost'' point extension operator (derived using boundary conditions)
so the concatenated equations of the system (as described above) can be written 
as a non-linear operator $\cE(\tilde\bx) = \bzero$ acting on the extended system variable,
where $\cE$ is the conservation equations operator (derived by concatenation of the system
equations).

The problem can be casted as to solve the following non-linear system:
\begin{eqnarray}
\bzero = \cE (\tilde{\bx}) = \cE (\cB (\bx)) = \cO(\bx)
\end{eqnarray}
 where $\cO = \cE \circ \cB$.

>>>>

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Since the system has axial symmetry, the operators are written in 
spherical coordinates $(r,\theta,\phi)$ 
(using the derivation at Appendix \ref{append:spherical}).

Due to axial symmetry, 
so the scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}
\end{eqnarray}
The scalar laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}
\end{eqnarray}
The vector gradient can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat
\end{eqnarray}

@@

\subsubsection{Grid Choice}
In order to discretize the differential operators, 
a regular grid of size $n_r \times n_\theta$ is used:
\begin{eqnarray}
(r_i,\theta_j) &\in& [1, \infty) \times [0,\pi] \\ 
r_i &=& (1+\Delta_r)^i \\
\theta_j &=& \Delta_\theta \cdot j
\end{eqnarray}
where a logarithmic grid is used for $r$ and an uniform one is used for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$ should be $\gg 1$. Thus:
\begin{eqnarray}
\Delta_r &=& \pars{R_{max}} ^ \frac{1}{n_r} - 1 \\
\Delta_\theta &=& \frac{\pi}{n_\theta}
\end{eqnarray}

@@

@@

The extended variables $\tilde X$ (with ``ghost'' points values 
from boundary conditions) are be computed, by applying the following non-linear operators:
\begin{eqnarray}
\tilde{\varPhi} &=& \cB_\varPhi(\varPhi, C; \beta) \\
\tilde{C} &=& \cB_C(C, \varPhi) \\
\tilde{V_r} &=& \cB_{V_r}(V_r; \cU) \\
\tilde{V_\theta} &=& \cB_{V_\theta}(V_\theta, C, \varPhi; \cU, \gamma) \\
\tilde{P} &=& P
\end{eqnarray}
The extended system variable can be written as $\tilde{\bx} = \cB(\bx)$, 
where $\cB$ is the ``ghost'' point extension operator (derived using boundary conditions)
so the concatenated equations of the system (as described above) can be written 
as a non-linear operator $\cE(\tilde\bx) = \bzero$ acting on the extended system variable 
$\tilde{\bx}$, where $\cE$ is the conservation equations operator 
(derived by concatenation of the system equations).

The problem can be casted to find a solution to the following non-linear system:
\begin{eqnarray}
\bzero = \cE (\tilde{\bx}) = \cE (\cB (\bx)) = \cO(\bx)
\end{eqnarray}
 where $\cO = \cE \circ \cB$.
